A Linear Solution of the Linearity Challenge

### 2025-12-21 paper/introduction.tex
cleanup
intro

The Concurrent Calculi Formalisation Benchmark~\citep{CarboneEtAl24} is a
collection of challenges concerning the mechanisation of formal models of
concurrent and distributed programming languages.
%
These models often make use of distinctive features that set them apart from the
formal models of \emph{sequential} programming languages, such as the adoption
of substructural (linear, affine) type systems, the dynamic scope of first-class
channels in systems of communicating processes, the need for coinductive
definitions and proof methods for describing and reasoning on possibly infinite
behaviors.
%
The benchmark aims at investigating effective formalization techniques that take
these features into account so as to foster the adoption of machine-checked
proofs in research work concerning concurrent and distributed programming
languages.

This paper reports on a solution to one of the challenges in the benchmark
concerned with the formalization of a minimal \emph{session calculus}. Hereafter
we will refer to this challenge as the \emph{linearity challenge}.
%
Sessions and session types~\citep{Honda93,HondaVasconcelosKubo98,HuttelEtAl16}
are established formalisms for the static analysis of distributed programs based
on peer-to-peer communication channels. Every session type system revolves
around three key ideas: (1) session endpoints are \emph{linear resources} that
cannot be discarded or duplicated without jeopardizing some safety and liveness
properties of a program; (2) the type of a session endpoint is \emph{updated}
after each use to reflect the state of the interaction it describes; (3) peer
session endpoints are meant to be used in complementary ways so as to guarantee
the absence of communication errors and, to some extent, progress of the
interaction.

The calculus that we formalize, which we call \Calculus, and the one in the
description of the challenge differ for two main aspects:
\begin{enumerate}
\item The challenge considers a minimal calculus of sessions, which is basically
    a first-order $\pi$-calculus without replication, whereas \Calculus is
    closely related to
    \CP~\citep{CairesPfenning10,Wadler14,CairesPfenningToninho16}, a
    fully-featured calculus of higher-order sessions with both linear and shared
    channels and whose type system is based on linear logic.
    %
    Targeting \Calculus allows us to consider a model of sessions that is
    well-known in the literature, that is based on firm logical grounds, and
    that includes a broader range of features including higher-order sessions
    and shared channels.
\item While the calculus in the challenge is equipped with native sessions,
    \Calculus is a \emph{linear} calculus where each channel is meant to be used
    for a \emph{single} communication. Structured conversations pertaining the
    same session and involving the exchange of several messages can still be
    modeled thanks to the use of \emph{explicit continuation channels}, as in
    the encoding of (binary) sessions into the linear
    $\pi$-calculus~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
    %
    Overall, \Calculus has the same expressive power of \CP but is a lower-level
    calculus in which sessions can be encoded instead of being featured
    natively.
\end{enumerate}

\Luca{Da qualche parte dire quali sono le sfide della sfida: representation of
channel names, context splitting. Parlare di co de bruijn?}

@@

The rest of the paper is structured as follows. \Cref{sec:calculus} describes
the syntax and the operational semantics of \Calculus and states the properties
of well-typed processes that we formalize and prove. \Cref{sec:agda} outlines
the essential aspects of the formalization with particular emphasis on the
representation of channel names and typing contexts. \Cref{sec:conclusion}
summarizes our contributions, discusses some extensions and future work. The
formalization has been carried out using Agda and its full code is available
from a public\Luca{Related work? Confronto con altre formalizzazioni?}
repository.\footnote{\url{https://github.com/boystrange/LinearityChallenge/tree/main/exp}}

>>>>

The Concurrent Calculi Formalisation Benchmark~\citep{CarboneEtAl24} is a
collection of challenges concerning the mechanisation of formal models of
concurrent and distributed programming languages.
%
These models often make use of distinctive features that set them apart from the
formal models of \emph{sequential} programming languages, such as the adoption
of substructural (linear, affine) type systems, the dynamic scope of first-class
channels in systems of communicating processes, the need for coinductive
definitions and proof methods for describing and reasoning on possibly infinite
behaviors.
%
The benchmark aims at identifying effective formalization techniques that take
these features into account so as to foster the adoption of machine-checked
proofs in research work concerning concurrent and distributed programming
languages.

One of the challenges in the benchmark, henceforth called \emph{linearity
challenge}, concerns the formalization of a minimal \emph{session calculus}.
%
Sessions and session types~\citep{Honda93,HondaVasconcelosKubo98,HuttelEtAl16}
are established formalisms for the static analysis of distributed programs based
on peer-to-peer communication channels. Every session type system revolves
around three key ideas: (1) session endpoints are \emph{linear resources} that
cannot be discarded or duplicated without compromising some safety and liveness
properties of a program; (2) the type of a session endpoint is \emph{updated}
after each use to reflect the state of the interaction it describes; (3) peer
session endpoints are meant to be used in complementary ways so as to guarantee
the absence of communication errors and, to some extent, progress of the
interaction.
%
The linearity challenge is motivated by the observation that the proper handling
of linear resources in a formalization often requires a large number of complex
auxiliary definitions and technical results that divert from the main problem
under consideration. One of the alleged culprits of such unwanted complexity is
\emph{context splitting}, namely the operation that partitions a typing context
in such a way that the linear resources described therein end up in only one of
the partitions. This observation has led to the development of various
techniques for dealing with typing contexts without splitting them, including
leftover typing~\cite{ZalakainDardha21}, the use of linearity
predicates~\cite{SanoKavanaghPientka23} and contexts with
tags~\cite{ZackonEtAl25}.

This paper presents a solution to the linearity challenge.
%
The calculus that we formalize, which we call \Calculus, and the one in the
challenge description differ for two key aspects.
%
First of all, the challenge only considers a minimal calculus of first-order
sessions, whereas \LCP is a fully-featured calculus that includes support for
higher-order, polymorphic sessions and both linear and shared channels. Also, as
suggested by its name, \LCP is closely related to \CP~\citep{Wadler14} and its
type system is based on classical linear logic.
%
The second difference is that the calculus in the challenge is equipped with
\emph{native} sessions, whereas \LCP is a \emph{linear} calculus where each
channel is meant to be used for a \emph{single} communication. Structured
sessions involving the exchange of several messages can be modeled thanks to the
use of \emph{explicit continuation channels}. This encoding of (binary) sessions
using linear channels is well known in the
literature~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
%
Overall, \LCP has the same expressive power of \CP, but is a lower-level
calculus in which sessions can be encoded instead of being featured natively.

The formalization we present is the most complete (in terms of features
supported by the calculus) and, at the same time, the simplest one (in terms of
code size). It is also one of only two formalizations that prove the deadlock
freedom property, and it does so with an order of magnitude less code than the
existing formalization~\cite{JacobsBalzerKrebbers22}.

the outcome of our investigation are somehow surprising in the sense that the
premises of the challenge and the assumption that the main technical challenge
is the representation of contexts

congetturiamo che la semplicità della formalizzazione derivi dalle
caratteristiche del calcolo più che dalle tecniche utilizzate, che alla fine
sono piuttosto standard.

@@

The rest of the paper is structured as follows. \Cref{sec:calculus} describes
the syntax and the operational semantics of \Calculus and states the properties
of well-typed processes that we formalize and prove. \Cref{sec:agda} outlines
the essential aspects of the formalization with particular emphasis on the
representation of channel names and typing contexts. \Cref{sec:conclusion}
summarizes our contributions, discusses some extensions and future work. The
formalization has been carried out using Agda and its full code is available
from a public\Luca{Related work? Confronto con altre formalizzazioni?}
repository.\footnote{\url{https://github.com/boystrange/LinearityChallenge/tree/main/cp}}
